### 2024-08-22 rebuttals/rebuttals.tex

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
The myopic approach (in pink) uses several instances of the pathway model, illustrated in Figure 1.1, as a sequence of different and linked LP problems. 
\end{mdframed}

\noindent and extended the caption of {\color{blue} Figure 1.4}. However, besides highlighting the key step 3.3 where there is an update of parameters value, I have decided to keep the figure as is since, on the one hand, the extension of the caption provides now a better understanding and, on the other hand, the code itself is exactly what is schematically represented on the diagram.

\begin{figure}[htbp!]
\centering
\includegraphics[width=10cm]{MY_process_code.pdf}
\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway. Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window. If the optimisation is not deterministic, step 3.3 is the one where the values of some parameters are changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
\label{fig:MY_process_code}
\end{figure}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{teal} \textbf{Stefan}} - With the myopic approach, are you predicting what will come or drawing a path that is the best? Is that a desirable feature to mimic the way we proceed now. Why artificially limit the model as we know we have to go to zero in 2050?
\end{mdframed}

\noindent As written in the manuscript and discussed during the private defense, the purpose of the myopic approach is to challenge the assumption of the complete knowledge about the uncertain future and mimic the short-sightedness of the decision-making process. This sequential optimisation allows showcasing robust policies to support myopic transitions (see Chapter 4) and the robustness of roadmaps (see Chapter 5). The myopic approach is also the methodological basis to support the implementation of the RL and PCA-based methods.

@@

\noindent Indeed, it is an effect of the salvage value as already written in the last paragraph ``Impact of myopic formulation on the system'' of Section 1.1: ``The main difference lies in the myopic transition itself and especially in the earlier
deployment of PVs and offshore wind turbines. These induce the reinforcement of the grid that is a capital-intensive and long-lifetime asset. This is mostly due to impact of the salvage value, Equation (1.4), in the objective function.''.

Sensitivity analysis has been performed on the formulation of the salvage value by \citet{goffauxpathway}. The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon. The methodology and results of this analysis is now summed up {\color{blue}in Appendix B.2}

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
In the work of \citet{goffauxpathway}, we carried out a sensitivity analysis on the formulation of the salvage value. First and foremost, this analysis confirmed the need to account for the salvage value to avoid penalising the capital intensive technologies towards the end of the transition \cite{poncelet2016myopic}. Then, the sensitivity analysis focused on two elements in the expression of the salvage value: $\tau_{phase}(p)$ and $\textbf{F\textsubscript{decom}}(p2,p,i)$.

In Equation (B.22), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built. With this expression, if a technology is still operational for half of its lifetime after 2050, then half of its initial investment is subtracted to the total transition cost. However, one could consider the depreciation of the asset where a technology that is still operational for half of its lifetime is worth less than half of its initial investment cost.

On top of this, we also investigated the impact of accounting for the decommissioned technologies or not in the expression of the salvage value.  In Equation (B.22), if a technology is decommissioned before 2050, then the salvage value of this technology will be 0. This is relevant since a technology decommissioned before 2050 would not be available after 2050. The issue is that the model could keep unnecessary technologies to subtract their salvage value from the total transition cost in order to decrease it. This would be done if the fixed OPEX of an unused technology is lower than its salvage value. Therefore, one could wonder if it would not be better to take into account the decommissioned technologies in the salvage value. In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a non-0 salvage value.

In a nutshell, the overall conclusion of this sensitivity analysis was that considering a salvage value was fundamental in our pathway formulation. However, the expression of the salvage value itself has little influence on the energy system at the end of the time horizon.  The interested reader is invited to refer to the work of \citet{goffauxpathway} for the numerical results of this analysis.
\end{mdframed}

@@

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
After assessing the sensitivity of the formulation of the salvage value (see Appendix B.2), we have kept the one given by Equation (1.4), being also similar to the expression used by \citet{prina2019transition}.
\end{mdframed}

\subsubsection{Discount rate and annualisation}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - Wouldn't it be more appropriate to talk about ``discount rate'' rather than ``interest rate''?
\end{mdframed}

\noindent Indeed, as we want to capture the time preference associated with obtaining finance for the project, effectively quantifying the trade-off between present and future financial value, it is more appropriate to refer to ``discount rate'' when considering the parameter $i_{\text{rate}}$. EnergyScope considers a state that needs to decide when investing in which technology. There is no consideration of return on investment as it would be the case for private investors. We have changed every occurrence of ``interest rate'' into ``discount rate'' {\color{blue}in the entire manuscript}.

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - Value of the discount rate is generally above 6\% in other works. You should have a better justification of selecting 1.5\%.
\end{mdframed}

\noindent Unlike snapshot models where the discount rate aims at annualizing investments, discounting in pathway models aims at evaluating the value of future expenses compared to present expenses. In other words, the higher the discount rate, the lower the impact of future extra costs. In their assessment of the role of the discount rates in energy systems optimisation models, \citet{garcia2016role} showed that higher ``social discount rate'', affecting the whole-energy system, including the operation (in opposition to technology-specific ``hurdle rates''), favours fossil-based technologies and postpones investments in more capital-intensive assets like solar PV and wind turbines. Where this is visible in the perfect foresight (PF) approach, the myopic optimisation of transition pathways is less affected by this given the shorter time-horizon considered in the sequential optimisations, 10 years versus 30 years in the PF approach. Even though the discount rate is usually between 7.5\% and 12\% in other studies, we have set its nominal value in line with G. Limpens' work \cite{limpens2021generating}. I have added some explanations in this regard {\color{blue}in Section 1.1.1}:

%\begin{figure}[htbp!]
%\centering
%\includegraphics[width=10cm]{tau_phase.png}
%\caption{$\tau_{phase}$ versus the timing of the expenses (columns) and the discount rate, $i_{rate}$ (rows).}
%\label{fig:tau_phase}
%\end{figure}

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
In EnergyScope Pathway, $\tau\textsubscript{\emph{phase}}$ aims at depreciating the value of expenses done in the future. The discount rate accounted for in this annualisation factor, $i_{\text{rate}}$, is considered as identical for all the technologies of the system. In line with \citet{limpens2021generating}, we set the nominal value to $i_{\text{rate}}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}. In the 30-year time window of the perfect foresight approach, having a higher value of discount rate would postpone investments in more capital-intensive assets. However, in a myopic vision, this effect is more limited given the shorter time window. 
\end{mdframed}

@@

Regarding the impact on the results, my educated guess is that this impact would be limited as the technologies of a same sector are assumed to have similar trends in terms of the evolution with time of their respective investment costs. Consequently, the major impact would be to end up with a lower total transition costs but the design of the system would be marginally impacted.

Given these further explanations, I have not brought any modification to manuscript in this regard.

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p.71 - Differentiated discount rates could be useful.
\end{mdframed}

\noindent As written in Chapter 3: ``In practice, the discount rate would vary depending on the technology investment risk''. Depending on the technology readiness level, private investors would be more or less risk-averse. However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capital, \ie private and public. Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capital. These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq. (1.3) when discussing the annualised phase factor}.

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system. In practice, the discount rate would vary depending on the technology investment risk. Depending on the technology readiness level, private investors would be more or less risk-averse. However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capital, \ie private and public. Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capital. For this reason, we have decided to keep an identical value of discount rate for all the technologies.
\end{mdframed}

@@

\subsubsection{Power grid and transport infrastructure}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - How do you model the power grid interconnections with the neighbouring countries as well as the transport network of the different energy carriers within the country? p.63 - The model decides to use ammonia instead of SNG $\rightarrow$ is the transport infrastructure accounted for? p.73 - How is the transport infrastructure accounted for?
\end{mdframed}

\noindent Regarding the power grid interconnections with the neighbouring countries, similarly to the import of other energy carriers, there is no representation of those physical routes. The whole-energy system of the country is represented as a single-node and the imported energy commodities are available ``at the front door'' of the system. It goes the same way for the infrastructures to transport energy carriers within the system. As it is represented assuming a copper-plate approach, all the demands, resources and conversion technologies are concentrated in one single node. This is now explicitly written {\color{blue}in a new paragraph of Section 2.2}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
Regarding the power grid interconnections with the neighbouring countries, similarly to the import of other energy carriers, there is no representation of those physical routes. The imported energy commodities are available ``at the front door'' of the system.  In terms of spatial resolution of the country itself, it is modelled as a single-node. This is similar to the copper plate model of a power grid. In other words, the infrastructures to transport the energy carriers within the country are not considered.  It is assumed that the demands have to be supplied by the production, without considering the flows between the producers and the consumers. Yet, adapting the networks is accounted for in terms of the required investments. For instance, a high share of VRES requires an investment to reinforce the power grid (\ie 368\,M€/GW of additional installed capacity of VRES). 
\end{mdframed}

@@

\noindent This comparison with a similar approach like factor fixing and factor prioritization (based on Morris method \cite{morris_factorial_1991}) is detailed {\color{blue}in the paragraph ``Comparison with a proven method'' of Section 1.2.2}. I have extended this paragraph by integrating the table of comparison between the Morris method and PCE.  This also avoids the reference to our own work and the reference [48], page 25 (see previous comment).

\noindent Besides being an in-house used method, the choice to use PCE is to be able to quickly extract statistical moments of the output of interest, as explained {\color{blue}in the first paragraph of Section 1.2.2}.

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p.21 - How are the groups of uncertainties defined? Is there a table?
\end{mdframed}

\noindent Even though the uncertain parameters, their groups and their ranges of uncertainties might be the same for other case studies, they have been especially adapted for the present case study, \ie Belgium. Consequently, I decided to exhaustively list these uncertain parameters in the chapter dedicated to the case study (see Section 2.4). However, to not leave the reader handing there and wondering if there are other parameters and how we have decided to group them, I have added a reference to Section 2.4 {\color{blue}at the end of Section 1.2.1} focusing on the characterisation of the uncertainties:

@@

\noindent The need of using NN is in the learning phase of the RL approach, not in running EnergyScope. The justification to use NN is already detailed in Section 1.3.2: ``Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of RL with Deep Neural Network (DNN), called Deep Reinforcement Learning (DRL). Among others, one of the main drawbacks of traditional RL algorithms, \ie without the use of Neural Network (NN),  is that it suffers from the ``curse of dimensionality'' when facing problems with continuous action and state spaces (see Chapter 4). By approximating the state-action function with its parameters (\ie weights and biases), DNN can address this difficulty. ''

Given what is already written in the manuscript, I have not brought modification in this regard.

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p. 27 When presenting the advantages of RL, you mention model-free approach. Ok, but you have a representation of the real world available to train the model!
\end{mdframed}

\noindent Indeed, we know the input parameters, the constraints and the objective function implemented in EnergyScope to optimise the transition pathway of the whole-energy system. In the RL framework, a model is a representation of how the environment will respond to the agent's actions. Unlike other cases such as grid world \cite{grid_world}, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.  Going for a model-free approach, we avoid building a model within the agent to inform it about this (complex) dynamics.

This aspect is already explained in Section 1.3.2: ``Given the assumed absence of knowledge of the agent about the dynamics of the environment, i.e., its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.'' One sentence of clarification has been added {\color{blue}in Section 1.3.2}:

\begin{mdframed}[style=manuscript] % Comment from the reviewer
In Reinforcement Learning, the ``model'' stands for the dynamics ``action-state-reward'' between the environment and the agent. 
\end{mdframed}

@@

\noindent As discussed during the private defense, it is more the spread of variation of the distribution that matters to characterise the robustness rather than the average position. This is now clarified {\color{blue}at the end of Section 1.4.2}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
It is really the range over which the distribution spans, $2\cdot \mathrm{MOE}$, that matters to characterise the variability of the projections of a roadmap. Indeed, as $\text{PC}_{\text{transition}}$ are vectors composed of positive and negative coefficients, the average position of the distribution, $\mu$, does not bring insight about the robustness of a roadmap.
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p. 39 Why are their outliers in the first case? Isn't a bit arbitrary to set them to a min/max value?
\end{mdframed}

\noindent As detailed in Section 1.4.2, even though these points result from the optimisation and make sense mathematically, they are defined as outliers as they fall out of the whiskers of their respective box plot. The way we managed these outliers is arbitrary but is a fair trade-off between leaving them as they are which would mislead the direction of PCs and completely removing them. As extensively discussed in Section 5.1.1, this pre-processing step will mostly affect the LT-heat sector where heat pumps running on methane are installed in rare cases, \ie less than 10\% of the 1260 transitions. Given the size of this sector, \ie between 22\% and 27\% of the EUD for Belgium, and the LP approach where there is a complete substitution of a technology by another one under certain conditions, these gas-HP can have a major weight in the system. However, to keep the main trends of variation without being ``polluted'' by outlying results (here, less than 10\% of the cases), we proceeded that way.

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{NED_tech.pdf}
\label{fig:NED_tech}
\end{figure}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p. 54 Is there a differentiation between ground mounted and rooftop PV?
\end{mdframed}

\noindent In the continuity of Gauthier Limpens' works \cite{limpens2021generating}, we consider here only rooftop PV. This potential of 59.2\,GW corresponds to 250 km$^2$ of available roof well oriented that exist today. Accounting for ground mounted PV would almost double this potential, as shown in Paolo Thiran's thesis. In future works, it would be interesting to also implement this additional potential, especially for the sake of harmonisation of input data. On the contrary, Elia, the Belgian TSO, foresees a much more limited increase of installed PV, \ie 18\,GW by 2034 \cite{Elia_2024_2034}. Consequently, for this thesis, I have decided to stick to this intermediate potential of 59.2\,GW. For the sake of clarity, I only added ``rooftop'' {\color{blue}in the second paragraph of Section 2.2}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
On the other hand, solar, wind, hydro and uranium are limited by the technical potentials of rooftop PV panels (59.2\,GW) [...]
\end{mdframed}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p. 51: Offshore wind max capacity is too low and PV is lower than Bregilab.
\end{mdframed}

\noindent Concerning the potential of PV, see previous comment. Regarding offshore wind, we have assumed a conservative potential of 6\,GW as the lower bound of the forecast done by the Belgian offshore platform\footnote{https://www.belgianoffshoreplatform.be/fr/}. Similarly to EnergyVille \cite{PATHS2050}, we could have gone up to the upper bound of 8\,GW. Going much higher, such as 24\,GW for the ``Electrification'' scenario of EnergyVille, would account for access to a part of the 150\,GW offshore wind park project in the North Sea signed between Belgium, Germany, Denmark and the Netherlands (the ``Esbjerg Offshore Wind Declaration'').

@@

\noindent As discussed during the private defense, we have decided to investigate to role of nuclear energy in the future for two main reasons: hot topic in Belgium and similar study by EnergyVille \cite{PATHS2050}. First, when presenting the results of our work at conferences/meetings, we have been asked many times how our results, massively relying on imports of renewable molecules, were influenced by the installation of nuclear power plants by 2050. Second, during my thesis, EnergyVille delivered a report PATHS2050 where, among others, they investigated the impact of SMR. As they observed that installing SMR would significantly reduce the need to import renewable molecules, we also wanted to assess this impact in our work. Given these explanations, no further modification has been brought to the manuscript in this regard.

### 2024-08-22 rebuttals/rebuttals.tex

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p. 53 4850€/kW is at the lower end of the EnergyVille scenario for SMR (4500€/kW). This is a very optimistic price forecast! Current tenders for 3rd generation are higher than 100€/MWh. There are papers showing that SMRs are not really expected to be cheaper than 3rd generation. How realistic is 41 €/MWh? LCOE of SMRs would be much higher if: - the capital cost was more realistic and, - the Capacity factor was lowered to account for their flexible use.
\end{mdframed}

\noindent As discussed during the private defense, similarly to EnergyVille \cite{PATHS2050}, we have considered the same CAPEX for SMR as the one considered for conventional nuclear power plants. Where EnergyVille considers a nominal CAPEX of 7500€/kW (like the current large Gen III design of Hinkley Point in the UK), we considered 4850€/kW like the CAPEX of conventional nuclear units in EnergyScope.

On top of the uncertainty applied to the CAPEX of SMR [-40\%, 44\%] considered in the thesis, I carried out a test on these assumptions by plotting the LCOE of electricity-production units considering different assumptions for SMR (see Figure \ref{fig:LCOE_comparison}). Even considering 50\% more expensive SMR subject to a higher discount rate (5\% versus 1.5\% for the other technologies), we notice that SMR keeps on being more cost-competitive than other flexible production units, like e-ammonia CCGTs, that are substituted by SMR when the latter can be installed (see Chapter 3). In other words, it means that the optimistic values assumed for SMR in the thesis lead to a cheaper whole-energy system but do not affect its design, \ie the installed technologies.  Consequently, no further modification has been brought to the thesis regarding this comment.

@@

To make it clearer from the beginning of the Chapter, I have extended {\color{blue}the contributions of Chapter 3}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
This GSA applied to a model optimising the transition pathway of a whole-energy system with these many uncertain parameters is the main contribution of this chapter. Through this analysis, we also managed to point out the impact of integrating SMR in the Belgian energy system as well as the main drivers of the import of e-hydrogen, e-methane, e-ammonia and e-methanol by 2050.
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - What is the learning of this chapter? I was underwhelmed by the long list of results. What is the take-away?
\end{mdframed}

\noindent The results are deeply detailed in Sections 3.1 and 3.2. These sections address the impact of adding SMR in the deterministic solution (\ie uncertain parameters at their respective nominal values) of the Belgian energy transition and global sensitivity analyses on different outputs of interest, respectively.  The purpose of these two sections is to give numerical substance to the main conclusions and take-aways that are listed in Section 3.3:
\begin{itemize}
\item If available, SMR is installed and mostly affect the electricity and high-temperature sectors;
\item To minimise the variation of the total transition cost, the key parameter is the cost of purchasing electrofuels where parameters related to SMR (availability and CAPEX) have much lower impact;
\item E-ammonia and, to a lesser extent, e-methane imports by 2050 are impacted by the availability of SMR whereas e-hydrogen and e-methanol are rather influenced by variation related to transport technologies and non-energy demand, respectively;
\item Even though SMR and electrofuels are in competition, they do not limit the deployment of VRES in Belgium. In other words, investing in wind and solar is a must-do for the Belgian energy transition;
\item Given the set of assumptions, especially considering the end-use-demands and the local potential of VRES, Belgium will need to import molecules in the near future. On top of investing into VRES, it seems reasonable to invest in the imports of these electrofuels as betting on SMR means letting the short-term emissions go up and, potentially, requiring significant (and quick) adjustments in case the technology is not ready on time to still respect the \ce{CO2} budget.
\end{itemize}

@@

\noindent There is indeed some curtailment, which remains very limited compared to the total production of solar-PV: 0.3\,TWh versus 61\,TWh. This information has been added {\color{blue}in the paragraph dedicated to ``High-temperature heat'' of Section 3.1.3}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
In the REF case, industrial electric heaters are mainly used to absorb the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days, and limit the yearly curtailment to a maximum of 0.3\,TWh, 0.5\% of the yearly 61\,TWh produced by solar-PV.
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - p. 69 - Very strange. Should be the same as for industrial boilers. No smart charging?
\end{mdframed}

\noindent The difference between BEV and industrial electric heaters is that BEV is already a preferred option for electrification, even in the REF case. BEV completely substitute diesel and gasoline-cars as soon as possible in the transition and there cannot be more cars as private mobility is limited to supply up to 50\% of the demand of passengers mobility. Unlike high-temperature heat, there is no modal swap due to the installation of SMR.

@@

\noindent Even though Reinforcement Learning can be applied to problems with infinite time-horizon discounting reward coming further in the future, it is not the case in my thesis. Indeed, each episode (\ie the transition) is finite. The purpose of Figure 4.1 was to focus on the reward function and how it articulates in the action-state-reward process. As already represented in this figure, three situations can terminate an episode: (i) infeasible optimisation, (ii) overshot \ce{CO2} budget before 2050 or, (iii) reaching 2050. This is now clarified {\color{blue}in the caption of Figure 4.1}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
Reward function, $R$. Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2} budget is exceeded. If the optimisation provides a solution and the \ce{CO2} budget is not exceeded, the episode continues. Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost, and the episode terminates. After terminating an episode, the process starts over at the initial state, \ie 2020.
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - How can you account for pivotal events/shocks in the system?
\end{mdframed}

\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/ shocks through uncertain parameters. For instance, this is how we have implemented the potential to install SMR in 2040 at the soonest. Given the probability described in Section 2.4, the technology can (or not) be installed. However, as detailed in Section 1.2.1, for the other uncertain parameters, they keep the same value relatively to their type and range for the whole transition. In other words, we do not allow zig-zag where, for instance, a cost of purchasing an energy carrier was low for the first half of the transition and, suddenly, in 2040, becomes very high.

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - p.91, Table 4.2: how do you differentiate between actions that depend on the agent and exogenous uncertainties? More in general: how is uncertainty integrated in the RL framework?
\end{mdframed}

\noindent Both, agent's actions and exogenous uncertainties, change the value of some parameters of EnergyScope Pathway. However, where the sample of uncertainties is drawn once and for all at the beginning of the episode as explained in Section 1.2.1, the agent's actions are taken at the beginning of each 10-year time window. This is now clarified {\color{blue}at the beginning of Section 1.3.2 and the updated version of Figure 1.7, and its caption} (see Figure \ref{fig:Schematics_RL} in the present document):

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
Before starting an episode, a sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section 1.2.1. 
\end{mdframed}

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{Schematics_RL.pdf}
\caption{The Reinforcement Learning (RL) framework applied to the myopic optimisation of the energy transition pathway between 2020 and 2050. Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years. At the beginning of each episode, a different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section 1.2.1.}
\label{fig:Schematics_RL}
\end{figure}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{teal} \textbf{Stefan}} - What is the potential of what we can learn from the results of the innovative methods like RL? We could go further in the analysis as we know we have to phase out coal, install renewables.
\end{mdframed}

\noindent First, I focus here below on each of the results sections of Chapter 4 highlighting the key conclusions already in the manuscript:

\begin{itemize}
\item \textbf{4.2.1 Reward and success}: (i) We have convergence of the learning process; (ii) Limiting success rate given the wide uncertainty ranges of some key parameters and the set of actions given to the agent; (iii) Near-term actions are needed.
\item \textbf{4.2.2 States}: (i) Validation of the formulation of the reward function (first respect the \ce{CO2} budget then minimise the cost); (ii) Reminder of the need for near-term actions; (iii) No-go zones in terms of share of renewables in the primary mix; (iv) Efficiency of the system does not give effective information for the agent to act upon.
\item \textbf{4.2.3 Actions}: (i) Limiting GWP and consumption of fossil gas are the keys; (ii) Even in a myopic perspective, limiting LFO is ineffective as it is phased out by the cost-driven model; (iii) Unlike coal that would still be part of the mix if the agent did not limit its consumption.
\item \textbf{4.3 Comparison with PF}: (i) Following \ce{CO2}-trajectory prescribed by PF would likely lead to failing the transitions with limited knowledge into the future; (ii) Succeeding myopic transitions relies more often on importing more electrofuels, mostly e-ammonia. 
\end{itemize}

Phasing out coal is, of course, necessary as it is a highly-polluting resource. However, what this work shows is that, whatever the stage in the transition, a policy limiting the use of coal will always be effective. However, to maximise the chance in succeeding the transition, the sooner the better. To insist on this aspect, this has been added {\color{blue}at the end of Section 4.2.3}:

\begin{mdframed}[style=manuscript] 
Whatever the stage in the transition, a policy limiting the use of coal will always be effective. However, to maximise the chance in succeeding the transition, the sooner the better.
\end{mdframed}

@@

Given these explanations and what is already in the text, I have not brought modifications to the manuscript in this regard.

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - p. 90: What does it mean to succeed/fail in 2030? And why are the number of attempts reduced going down in the graph?
\end{mdframed}

\noindent As discussed during the private defense, the grey curves represent the distribution of occurrences failing (\ie infeasible optimisation or overshooting \ce{CO2} budget) by 2050 at the latest. The number of remaining attempts reduce as prematurely ended episode are removed from the further counts. The purpose of displaying these numbers of ``Total attempts'' is to explain why the areas below the orange curves increase relatively to the grey curves. These two aspects are now explicitly explained {\color{blue}in the caption of Figure 4.4 and the other similar figures}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right). \textbf{Number of remaining attempts decrease with time since infeasible problems and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050. Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2} budget and are cheaper than the REF case. 
\end{mdframed}

@@

\noindent In the context of Reinforcement Learning, a step consists in taking an action, collecting a reward based on the previous state and the taken action and ending in a new state. Even though it was already mentioned in Chapter 1, the fact that the results come further from the methodology part makes it harder for the reader to remember the meaning of a step. It is now more explicitly defined {\color{blue}in the first paragraph of Section 1.3.2}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
From this state, the agent takes a step: action, reward, new state.
\end{mdframed}

\noindent and reminded {\color{blue}in the first sentence of Section 4.2.1}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
The learning phase has been split into batches of 500 steps, \ie 500 sequences of state-action-reward-new state. 
\end{mdframed}
